### 2019-02-10 concepts.tex
fix n to i
fix
more text, fix formalisms

\section{Practical Lightweight Nondeterminism
  Detection and Debugging}

@@

We assume that the underlying behavior of a system may be
deterministic, or nondeterministic, in the strict sense, by allowing
for a transition relation that \emph{may} be a function of
$S \times A$.  A \emph{trace} $t$ is a finite sequence $t = s_0
\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} s_n$ where $s_0 \in I$ and $\forall n
. (s_n, \alpha_n, s_{n+1}) \in T$.

@@

To define failure nondeterminism, we extend the
transition relation to include a notion of \emph{failure}: $T \subseteq S
\times A \times S \times {\tt bool}$, where the boolean indicates
whether the action \emph{failed}.  A trace $t = s_0
\xrightarrow[]{\alpha_0} (s_1, F_0) \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} (s_n, F_n)$ shows \emph{failure
  nondeterminism} if $\exists i$ such that $F_{i+1}$ (that is,
$\alpha_i$ \emph{fails}) and either (1) $v(s_{i}) \neq v(s_{i+1})$ or (2)
$\alpha_i = \alpha_{i+1} \wedge \neg F_{i+2}$.  The second possibility
captures our more lightweight approach to checking for failure nondeterminism.

\subsection{Nondeterminism and Delta-Debugging}

@@

\subsubsection{``Publication Bias'' in Delta-Debugging}
\label{sec:pubbias}

@@

>>>>

\section{Practical Lightweight Nondeterminism
  Detection}

@@

We assume that the underlying behavior of a system may be
deterministic, or nondeterministic, in the strict sense, by allowing
for a transition relation that \emph{may} be a function of
$S \times A$.  A \emph{trace} $t$ is a finite sequence $t = s_0
\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} s_n$ where $s_0 \in I$ and $\forall i < n
. (s_i, \alpha_i, s_{i+1}) \in T$.

@@

To define failure nondeterminism, we extend the
transition relation to include a notion of \emph{failure}: $T \subseteq S
\times A \times S \times {\tt bool}$, where the boolean indicates
whether the action \emph{failed}.  A trace $t = s_0
\xrightarrow[]{\alpha_0} (s_1, F_0) \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} (s_n, F_n)$ shows \emph{failure
  nondeterminism} if $\exists i$ such that $F_{i+1}$ (that is,
$\alpha_i$ \emph{fails}) and either (1) $v(s_{i}) \neq v(s_{i+1})$ or (2)
$\alpha_i = \alpha_{i+1} \wedge \neg F_{i+2}$.  The second possibility
captures our more lightweight approach to checking for failure
nondeterminism.  Note that unlike all of the other ways to demonstrate
nondeterminism, this second type of failure nondeterminism does not
require the use of $v$, or even the notion of observable components of
a state, or (for that matter) any comparison of states.

A further point to emphasize is that while horizontal determinism
requires two traces, it does not require a test generation algorithm
to produce pairs of traces, which would require substantial
modification of most such tools (e.g., random testers or fuzzers, and
even model checkers).  In general, testing tools do not produce
\emph{traces}, they produce \emph{action sequences}.  The two traces
used to show nondeterminism, by definition, have the same action
sequences, which means that a testing tool only needs to search for an
action sequence which, when executed twice, produces a $t_1$ and $t_2$
meeting the requirements for showing nondeterminism ---
non-equivalence of $v(s)$ for some $s$.

\section{Nondeterminism and Delta-Debugging}

@@

\subsection{``Publication Bias'' in Delta-Debugging}
\label{sec:pubbias}

@@

\subsection{Using Replications to Mitigate ``Publication Bias''}

### 2019-02-10 implementation.tex
more text, fix formalisms

In order to implement a practical nondeterminism-detection tool, it is
important to consider the sources of nondeterminism.  Most of these
sources can, in principle, be detected simply by re-running a test
twice and comparing results, either at all visible values or at the
final state.  For example, if some library call uses a random number
generator (perhaps in a stochastic algorithm), the value should be
different on a second run.  Timing dependencies may change on a second
execution, especially if a delay is added between steps of the test (a
more principled version of the unfortunate tradition of testing
concurrent code by adding ad-hoc {\tt sleep} statements).  However,
simply running a test twice (possibly with a delay between steps) does
not suffice in all cases.

>>>>

In order to implement a practical nondeterminism-detection tool, it is
important to consider the sources of nondeterminism.  Most of these
sources can, in principle, be detected simply by re-running a test
twice and comparing results, either at all visible values or at the
final state.  For example, if some library call uses a random number
generator (perhaps in a stochastic algorithm), the value should be
different on a second run.  Timing dependencies may change on a second
execution, especially if a delay is added between steps of the test (a
more principled version of the unfortunate tradition of testing
concurrent code by adding ad-hoc {\tt sleep} statements).   Most
nondeterministic behavior in tests is probably attributable to a such
effects that can vary with a repeated execution, in the same
(approximate) enviroment, of the same test.  The two broad classes of
nondeterminism that can usually be detected (with some probability) by
simply executing a test twice in succession are:

\begin{itemize}
\item {\bf External Environment:} This class of nondeterminism arises
  when a test's behavior depends on factors outside the program under
  test, which are subject to uncontrolled variance.  Calling {\tt
    random()} in Python is an obvious example; the call is almost
  guaranteed to return a different value each time it is called in a
  test, unless we explicitly re-seed the random number generator at
  the beginning of each test.  Similarly, calls to {\tt time}
  functions  will usually return different values on different runs,
  and even if only elapsed time, rather than absolute, is used to
  influence behavior, subtle variances in timing are to be expected
  due to system load, I/O, and other external factors.  Finally, in
  distributed system testing, the environment is often remote servers
  (or clients) that may be unavailable, slower than expected, or
  subject to faulty behavior for reasons not under the control of the test.
\item {\bf Concurrency:} Perhaps the second most common cause for
  nondeterminism in tests is when an algorithm is implemented using
  concurrency, in the (incorrect) beleif that the result of the
  algorithm remains deterministic.  E.g., consider a multi-threaded
  implementation of a search algorithm, that returns the location of
  an item in an unsorted list.  If the list may contain duplicates,
  and a developer has not accounted for this possibility, the
  algorithm may return different locations of an item for the same list,
  depending on the scheduling of threads.  Arguably, this can be
  subsumed under the first class, with the behavior of the
  processor/scheduler as the ``enviromental'' factor; relying on
  scheduling is, in a sense, equivalent to calling {\tt random} or
  {\tt time}.
\end{itemize}

However, simply running a test twice in the same process (possibly with a delay between steps) does
not suffice in all cases.
